A High-throughput, Scalable Solution for the New York Taxi Trips DEBS Grand Challenge Problem

### 2015-04-28 paper/conclustion.tex
commiting last changes

\section{Conclusion}

We propose a scalable solution for DEBS grand challenge problem 2015. Our queries execute with O(log(n)) time complexity and hence scales with window size. We propose a scalable evaluation technique to execute queries in a distributed setup. Since \textit{EventEmitter} partitions the events using the  hash code value of the key (\textit{Route} or \textit{Cell}), our solution can be used to process events with multiple \textit{EventEmitters} (multiple event streams) as well. Different levels of \textit{ResultAggregators} can be used to scales up aggregation process if required. We propose an algorithm to find the runtime median with O(log(n)) time complexity. This algorithm specially suites for finding median with repeating values. The time windows used in our solution has an O(1) time complexity but O(n) space complexity. This creates opportunity to research further on efficient memory management techniques for large window sizes.

>>>>

\section{Conclusions and Future Work}

We propose a scalable solution for DEBS grand challenge problem 2015. Our queries execute with O(log n) time complexity and hence scale with window size. We propose a scalable evaluation technique to execute queries in a distributed setup. Since \textit{EventEmitter} partitions the events using the  hash code value of the key (\textit{Route} or \textit{Cell}), our solution can be used to process events with multiple \textit{EventEmitters} (multiple event streams) as well. Different levels of \textit{ResultAggregators} can be used to scale up aggregation process if required. We propose an algorithm to find the runtime median with O(log n) time complexity. This algorithm is especially suited for finding median with repeating values. The time windows used in our solution have an O(1) time complexity, but O(n) space complexity. This creates opportunities for future research on efficient memory management techniques for large window sizes.

### 2015-04-28 paper/experiments.tex
commiting last changes

\section{Experiments}
We conducted several experiments to study the scalability of our solution. Firstly we study the throughput and mean message latency variation with the number of \textit{QueryEvaluators} for both single JVM and distributed setups to find the scalability of the parallel evaluation. We measured the end to end latency. At the \textit{EventEmitter}, we set the time stamp for all events once the record is read. At the \textit{QueryEvaluator} we copy the original time stamp, if that event causes a top ten value change event. Total delay calculated at the \textit{ResultAggregator} just before writing the event. In an ideal system, we should observe linear throughput increment with a constant message latency. However there are some exceptions and we have provided our explanations on them. Finally we studied the throughput variation with the window size to study the suitability of our solution for higher window sizes. All these experiments were performed on a LAN with a network bandwidth of 1Gbps and all machines were Intel(R) Xeon(R) 2.4GHz, 4 core duo machines with 16 GB of memory.

\subsection{Frequent routes query}

@@

We deployed this process graph to a distributed stream processing system we have developed, and  measured the throughput and the mean message latency varying the number of \textit{RouteProcessors} with the data set given (173 million records). In the local set up, we use a new thread to execute a new instance and in the distributed setup we add a new node to execute the new instance.

@@

Figure \ref{latency_route} shows the latency variation. For single JVM setup latency get reduced at 2 and slightly increases after that due to context switching overhead with higher number of threads. For distributed setup there is an high initial latency due to communication overheads. After that it comes to a constant value.

\subsection{Profitable cells query}

@@

Figure \ref{profictgraph} shows the process graph for profitable cells query. \textit{ProfitEventEmitter} reads the data file and calculates the pickup cell and dropoff cell details using longitude and latitude coordinates. Then it creates two events one for dropoff part and other to pickup part, with different sequence numbers and distributes the events among \textit{ProfitCalculators} using the hash value of the cell. In this way cells are partitioned into different \textit{ProfitCalculator} instances and each instance gets all the events relevant to its' cell set. We use a set of threads (equals to the number of Profict Calculators) to emit data as in the previous query. Finally \textit{TopProfitProcessor} aggregates events and writes to a file.

@@

The grand challenge problem uses 15 minute window size for profitable cells query. However in a real world problems window sizes may required. Therefore finally we evaluated scalability of our solution with the window size.  For this experiment, we measured the throughput of the profitable cells query using a single JVM and three instances by varying the window size. Further we compared the throughput with our \textit{DynamicHeap} based solution which is not effective in lower window sizes as well.

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{window.png}
        \caption{Throughput variation with the window size.}
        \label{window}
\end{figure}

Figure \ref{window} shows the results. For both solutions, throughput does not drop by half, even though the window size increases 64 times. And for large windows \textit{DynamicHeap} performs better than the \textit{PriorityQueue} based one. This is due to two reasons. Firstly, all operations of \textit{DynamicHeap} are O(log(n)). Secondly, out of 173 million records there are only around 18000 unique fares. Therefore \textit{DyamicHeap} based algorithm does not consume memory beyond that. However , since \textit{PriortyQueue} based implementation keeps all values in memory it consumes more memory for large windows.

>>>>

\section{Experiments}
We conducted several experiments to study the scalability of our solution. We profiled the throughput and mean message latency variation with the number of \textit{QueryEvaluators} for both single JVM and distributed setups to find the scalability of the parallel evaluation. We measured the end to end latency. At the \textit{EventEmitter}, we set the time stamp for all events once the record is read. At the \textit{QueryEvaluator} we copy the original time stamp, if that event causes a top ten value change event. Total delay is calculated at the \textit{ResultAggregator} just before writing the event. In an ideal system, we should observe linear throughput increment with a constant message latency. However there are some exceptions and we have provided our explanations on them. Finally we studied the throughput variation with the window size to study the suitability of our solution for higher window sizes. All these experiments were performed on a LAN with a network bandwidth of 1Gbps and all machines were Intel(R) Xeon(R) 2.4GHz, 4 core duo machines with 16 GB of memory.

\subsection{Frequent Routes Query}

@@

We deployed this process graph to a distributed stream processing system we have developed, and  measured the throughput and the mean message latency varying the number of \textit{RouteProcessors} with the dataset given (173 million records). In the local set up, we use a new thread to execute a new instance and in the distributed setup we add a new node to execute the new instance.

@@

Figure \ref{latency_route} shows the latency variation. For single JVM setup latency get reduced at 2 and slightly increases after that due to context switching overhead with higher number of threads. For distributed setup there is a high initial latency due to communication overheads. After that it comes to a constant value.

\subsection{Profitable Cells Query}

@@

Figure \ref{profictgraph} shows the process graph for profitable cells query. \textit{ProfitEventEmitter} reads the data file and calculates the pickup cell and dropoff cell details using longitude and latitude coordinates. Then it creates two events one for dropoff part and other to pickup part, with different sequence numbers and distributes the events among \textit{ProfitCalculators} using the hash value of the cell. In this way cells are partitioned into different \textit{ProfitCalculator} instances and each instance gets all the events relevant to its' cell set. We use a set of threads (equals to the number of \textit{ProfitCalculators}) to emit data as in the previous query. Finally \textit{TopProfitProcessor} aggregates events and writes to a file.

@@

The grand challenge problem uses 15 minute window size for profitable cells query. However in real world problems higher window sizes may required. Therefore finally we evaluated scalability of our solution with the window size.  For this experiment, we measured the throughput of the profitable cells query using a single JVM and three instances by varying the window size. Further we compared the throughput with our \textit{DynamicHeap} based solution which is not effective in lower window sizes as well.

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{window.png}
        \caption{Throughput variation with window size.}
        \label{window}
\end{figure}

Figure \ref{window} shows the results. For both solutions, throughput does not drop by half, even though the window size increases 64 times. For larger windows \textit{DynamicHeap} performs better than the \textit{PriorityQueue} based one. This is due to two reasons. First, all operations of \textit{DynamicHeap} are O(log n). Second, out of the 173 million records there are only around 18000 unique fares. Therefore the \textit{DyamicHeap} based algorithm does not consume memory beyond that. However, since \textit{PriortyQueue} based implementation keeps all values in memory, it consumes more memory for larger windows.

### 2015-04-28 paper/introduction.tex
commiting last changes

The Grand Challenge 2015 problem focuses on processing New York city taxi trip data for year 2013. The area considered is a square of 150m x  150m. The  square is divided into smaller units called cells and a taxi trip is considered as moving from one cell (pickup cell) to another cell (dropoff cell). Therefore a route is uniquely identified by the pickup cell and the dropoff cell. The data set consists of 173 million events and solutions are expected to replay these data and generate events that meet criteria of two queries defined.

The first query is to find the events which change the top ten routes within last 30 minutes and emit the new route set details along the original event. The given data set contains 899010  routes and the 30 minute time window for this query accounts around 10000 events assuming an equal event distribution. This creates the possibility of processing several thousand route counts specially within peak hours. In other words we need to evaluate several thousand route counts with each event and find 10 top most routes. Similar analysis can be made with profitable cells query as well. In addition to that it requires to  calculate the median fare for each cell.

\subsection{Research Challengers}

If we consider an application to process this 173 million data set within an year, it would be a trivial task and any naive implementation satisfy that requirement. However if we consider an application which process these data within minutes with minimal latency to process each event, we observe following research challengers.
\begin{enumerate}
	\item \textit{Handling thousands of values in real time :}  An algorithm with O(n) time complexity will not scale for thousands. Therefore is it possible to come up with an algorithm of O(log(n)) time complexity?
	\item \textit{Parallel evaluation :}  This application is not inherently support parallel evaluation. Partitioning route sets and cell sets triggers events for sub sets, but application requires the events with whole sets. Therefore is it possible to partition data and still generate required results to support parallelism?  
\end{enumerate}

@@

We started our work seeking answers for above challengers. First we evaluated doubly linked list and heap based approaches to build a data structure with O(log(n)) time complexity. After several evaluations, we developed a new data structure called \textit{NodeList} which utilizes both concepts. In order to find the running median with O(log(n)) time complexity, we experimented with an algorithm which uses priority queue and the heap data structure we developed. We decided to use former since it gave better performance with smaller window sizes. Then we focused on parallelizing the application and came up with a strategy to aggregate the sub set results to find the final results. As the last step, we conducted several experiments to evaluate the scalability of our solution. We repeated this process identifying possible issues and fixing them.

\subsection{Contributions}
We observe following key contributions, which can be used in solving future DEBS challenges as well.
\begin{enumerate}
	\item We have developed a set of  generic data structures capable of inserting, updating, removing and retrieving top values (order can be defined using a comparator function) with O(log(n)) time complexity. 
	\item Using our data structures we have developed a new algorithm to find the running median with O(log(n)) time complexity. One of the key challenges of DEBS 2014 \cite{jerzak2014debs}  was to find the median of a large time window. We believe our solution is useful in such situations.
	\item We have developed a scalable parallel evaluation technique to process queries efficiently. 
\end{enumerate}

\subsection{Paper Organization}
The remainder of this paper is organized as follows. Section 2 provides an in-depth description about the internal data structures and the algorithms used to process the events. Section 3 describes our scalable design to evaluate queries and associated problems. Section 4 proves the scalability of our solution with experimental results. Section 5 discuss the related work including some last year solutions. Section 6 summarizes key concepts and discuss their further applications.

>>>>

The Grand Challenge 2015 problem focuses on processing New York city taxi trip data for the year 2013. The area considered is a square of 150km x  150km. The  square is divided into smaller units called cells and a taxi trip is considered as moving from one cell (pickup cell) to another cell (dropoff cell). A route is uniquely identified by the pickup cell and the dropoff cell. The dataset consists of 173 million events and solutions are expected to replay these data and generate events that meet the criteria of two defined queries.

The first query is to find the events that change the top ten routes within last 30 minutes and emit the new route set details along the original event. The given dataset contains 899010 routes and the 30 minute time window for this query accounts around 10000 events assuming an equal event distribution. This creates the possibility of processing several thousand route counts especially during peak hours. In other words we need to evaluate several thousand route counts with each event and find 10 top routes. A similar analysis can be made for the profitable cells query as well. In addition to that it requires to calculate the median fare for each cell.

\subsection{Research Challenges}

If we consider an application that must process the 173 million trips dataset within an year, it would be a trivial task and several simple implementations would satisfy that requirement. However if we consider an application that must process these data within minutes with the added constraint of minimal latency to process each event, we observe following research challenges.
\begin{enumerate}
	\item \textit{Handling thousands of values in real time :}  An algorithm with O(n) time complexity will not scale for thousands. How can we design an algorithm with O(log n) time complexity?
	\item \textit{Parallel evaluation :}  This application does not inherently support parallel evaluation. Partitioning route sets and cell sets triggers events for subsets, but application requires the events with whole sets. Therefore is it possible to partition data and still generate required results to support parallelism?  
\end{enumerate}

@@

We started our work seeking answers for the above challenges. First we evaluated doubly linked list and heap based approaches to build a data structure with O(log n) time complexity. After several evaluations, we developed a new data structure called \textit{NodeList} that utilizes both these concepts. In order to find the running median with O(log n) time complexity, we experimented with an algorithm which uses priority queue and the heap data structure we developed. We decided to use former since it gave better performance with smaller window sizes. We then focused on parallelizing the application and designed a strategy to aggregate the subset results to find the final results. We also conducted several experiments to evaluate the scalability of our solution.

\subsection{Contributions}
We observe following key contributions, which can be used in solving future DEBS challenges as well.
\begin{enumerate}
	\item We have designed a set of  generic data structures capable of inserting, updating, removing and retrieving top values (order can be defined using a comparator function) with O(log n) time complexity. 
	\item Using our data structures we have developed an algorithm to find the running median with O(log n) time complexity. One of the key challenges of DEBS 2014 \cite{jerzak2014debs}  was to find the median of a large time window. We believe our solution is useful in such situations.
	\item We have developed a scalable parallel evaluation technique to process queries efficiently. 
\end{enumerate}

\subsection{Paper Organization}
The remainder of this paper is organized as follows. Section 2 provides an in-depth description about the internal data structures and the algorithms used to process the events. Section 3 describes our scalable design to evaluate queries and associated problems. Section 4 proves the scalability of our solution with experimental results. Section 5 discusses the related work including some last year solutions. Section 6 summarizes key concepts and discusses their further applications.

### 2015-04-28 paper/paper.tex
commiting last changes

\maketitle
\begin{abstract}
Processing complex queries on unbounded event streams in real time, is a challenge for many organizational data processing systems. These systems are expected to process data with reduced latency to generate real time events, and high throughput to minimize the required hardware. In this regard, DEBS Grand Challenge 2015 focuses on evaluating two queries (frequent routes and  profitable cells) in real time with low latency and high throughput. These queries involve processing windows of thousands of records. Firstly, such processing demands efficient data structures and algorithms to minimize the processing overhead. Secondly, the system should partition data to evaluate them in parallel to make it scalable.

In this paper, we present a set of data structures which can be used to evaluate the aforementioned queries with O(log(n)) time complexity and a data partitioning technique to evaluate them in parallel. Then we evaluate our solution in a signal machine as well as in a distributed setting in a commodity cluster of machines over a 1Gbps LAN. We could process frequent routes query with 173 million data set within 5 minutes with less than 4 millisecond latency and profitable cells query with same data set within 11 minutes with less than 5 millisecond latency.
\end{abstract}

>>>>

\maketitle
\begin{abstract}
Processing complex queries on unbounded event streams in real time, is a challenge for many data processing systems. These systems are expected to process data with reduced latency to generate real time-events, and at high throughput to minimize the required hardware. In this regard, DEBS Grand Challenge 2015 focuses on evaluating two queries (frequent routes and  profitable cells) in real time with low latency and high throughput. These queries involve processing windows of thousands of records. Firstly, such processing demands efficient data structures and algorithms to minimize the processing overhead. Secondly, the system should partition data to evaluate them in parallel to make it scalable.

In this paper, we present a set of data structures that we designed to evaluate the aforementioned queries with O(log n) time complexity and a data partitioning technique to evaluate them in parallel. We then evaluate our solution on a signal machine as well as in a distributed setting in a commodity cluster of machines over a 1Gbps LAN. We are able to process the frequent routes query with the 173 million trips dataset within 5 minutes with less than 4 millisecond latency and the profitable cells query with same dataset within 11 minutes with less than 5 millisecond latency.
\end{abstract}

### 2015-04-28 paper/query_evaluation.tex
commiting last changes

\subsection{Sequential evaluation}

@@

First both processors can be executed in same JVM with same thread. In that case, both processes execute sequentially. We can make two processes parallel by executing them with two threads and communicate using a queue in same JVM or communicate using a TCP connection for different JVMs. In all these cases \textit{QueryEvaluator} processes messages sequentially and becomes the bottleneck. Next we show how to parallel \textit{QueryEvaluator} process.

\subsection{Parallel evaluation}
The general way to parallelize event processing, is to partition the events according to a key and run multiple instances to process data. Therefore in this section, we explain how to parallelize the query evaluation process taking frequent route query into consideration. Same concepts can be applied to profitable cell evaluation as well.

@@

Algorithm \ref{aggregate_algorithm} shows the algorithm for ResultAggregator with above fixes. First it removes the expired keys. Then it updates the existing values or adds new values and moves the position in the list accordingly. Finally it generates an event if there is a change to the top list.

>>>>

\subsection{Sequential Evaluation}

@@

First both processes can be executed in same JVM with same thread. In that case, both processes execute sequentially. We can make two processes parallel by executing them with two threads and communicate using a queue in same JVM or communicate using a TCP connection for different JVMs. In all these cases \textit{QueryEvaluator} processes messages sequentially and becomes the bottleneck. Next we show how to parallel \textit{QueryEvaluator} process.

\subsection{Parallel Evaluation}
The general way to parallelize event processing, is to partition the events according to a key and run multiple instances to process data. Therefore in this section, we explain how to parallelize the query evaluation process taking frequent route query into consideration. Same concepts can be applied to profitable cell evaluation as well.

@@

Algorithm \ref{aggregate_algorithm} depicts the algorithm for \textit{ResultAggregator} with the above fixes. The algorithm first removes the expired keys. Then it updates the existing values or adds new values and moves the position in the list accordingly. Finally it generates an event if there is a change to the top list.

### 2015-04-28 paper/query_implementation.tex
commiting last changes

\section{Query Implementation}
In this section, we elaborate how we have implemented the query logic to achieve O(log(n)) time complexity. The 2015 Grand Challenge problem has two queries. First is to evaluate top ten most frequent routes within the last 30 minutes and second is to evaluate top ten profitable areas within the last 15 minutes. We can use composite objects to manage details for each route and cell and compare them to find top 10 values. However, each time window can have thousands of route and cell details. Therefore, an efficient data structure is required to add, remove, update and retrieve top ten values from a set of objects for a given time window efficiently. Further, since each event is received with route and cell details, the aforementioned operations should be performed using route and cell details as the key. In order to support all these functions with O(log(n)) time complexity we have developed a data structure called \textit{NodeList}. The remaining of this section describes how we have achieved this time complexity with the \textit{NodeList} data structure and how we have implemented the queries using that data structure.

\subsection{NodeList data structure}

\begin{table}
\centering
\caption{Operations of \textit{NodeList} Data Structure}
\begin{tabular}{|l|l|} \hline
Operation Signature & Time Complexity \\ \hline \hline
add(key : Object, value : NodeValue) & O(log(n)) \\ \hline 
get(key : Object) : NodeValue & O(1) \\ \hline
remove(key : Object) : NodeValue & O(log(n)) \\ \hline
decrementPosition(key : Object) & O(log(n)) \\ \hline
incrementPosition(key : Object) & O(log(n)) \\ \hline
getTopValues() : List<NodeValue> & O(1) \\ \hline
\end{tabular}
\label{nodelist_api}
\end{table}

Table \ref{nodelist_api} shows the operations of this data structure. The \textit{NodeValue} is an interface to plug any user defined type as a value. For an example, the route query can use a \textit{RouteCount} object to keep route count while profitable cells query can use \textit{CellProfit} object to keep track of fare details. This interface contains a method called \texttt{compare} to define the order of the values. The \textit{NodeList} data structure provides the notion of a sorted values starting from position 0 (highest value) to its users. The \texttt{decrementPostion} operation moves the corresponding value towards position 0 until the correct position according to sorted order. Therefore this method should be called after increasing the value of an existing object. Similarly \texttt{incrementPosition} operation can be used to move a value to correct position after decreasing the value of an existing object. \texttt{GetTopValues} operation returns the top ten values of the list. The query algorithm sections further elaborate usage of these methods.

One way of implementing this data structure is to use a heap. A heap supports adding and extracting maximum value operations in O(log(n)) time. However since this application requires top 10 values, we need to retrieve maximum value ten times and insert them back. On the other hand if we use a doubly linked list to keep values, it will take O(n) time for  \texttt{decrementPostion} and \texttt{incrementPosition} operations. Therefore we can use a doubly linked list to keep first 10 values (this makes time complexity for all operations with this data structure O(1)) and a heap to keep other values to avoid above problems.  A \textit{Map} which points to values can be used to retrieve values using a key.

@@

\subsection{OrderedList data structure}

@@

Table \ref{orderedlist_api} shows the operations of the \textit{OrderedList} data structure. First six methods has the same meaning as the \textit{NodeList} data structure and last two can be used to get the size of the list and the value of the last object. We internally keep a counter and a pointer to tail to support these operations in O(1) time. Figure \ref{orderedlist_impl} shows the implementation details. The \textit{Map} structure (an \textit{java.util.HashMap}) is used to retrieve object pointers with O(1) expected time. Each node of the doubly linked list points to previous and next nodes while storing the \textit{NodeValue}.

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{orderedlist.png}
        \caption{\textit{OrderedList} data structure implementation}
        \label{orderedlist_impl}
\end{figure}

\subsection{DynamicHeap data structure}

\begin{table}
\centering
\caption{Operations of \textit{DynamicHeap} Data Structure}
\begin{tabular}{|l|l|} \hline
Operation Signature & Time Complexity \\ \hline \hline
add(key : Object, value : NodeValue) & O(log(n)) \\ \hline
remove(key : Object) : NodeValue & O(log(n)) \\ \hline
containsKey(key : Object) : boolean & O(1) \\ \hline
get(key : Object) : NodeValue & O(1) \\ \hline
moveUp(key : Object) & O(log(n)) \\ \hline
moveDown(key : Object) & O(log(n)) \\ \hline
getMaxkey() : Object & O(1) \\ \hline
extractMax() : NodeValue & O(log(n)) \\ \hline
\end{tabular}
\label{dynamicheap_api}
\end{table}

Table \ref{dynamicheap_api} shows the operations of the \textit{DynamicHeap} data structure. \texttt{MoveUp} and \texttt{moveDown} operations are used to move an element up in the heap if the value increased, or move down in the heap if the value decreased. Figure \ref{dynamicheap_impl} shows the implementation details. Conventional heaps only support \texttt{add} and \texttt{extractMax} operations since heap operations require knowing the position of the element in underlying array. In order to support \texttt{moveUp}, \texttt{moveDown} and \texttt{remove} operations, we keep the array index with the object itself. As shown in figure \ref{dynamicheap_impl} \textit{HeapNode} object contains the array index and the \textit{NodeValue} given by the user. This allows us to retrieve array index in O(1) time and move the element O(log(n)) time.

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{DinamicHeap.png}
        \caption{\textit{DynamicHeap} data structure implementation}
        \label{dynamicheap_impl}
\end{figure}

\subsection{Finding Runing Median}
In general, we can find runing median using two heaps. One to keep values greater than the current median and one to keep values less than the current median. An implement of this algorithm using two \textit{java.util.PriorityQueue}s have an O(log(n)) time complexity for add operation and O(n) time complexity for remove operation (\texttt{remove} operation of \textit{java.util.PriorityQueue} takes O(n) time).  We overcame this issue using \textit{DynamicHeap} to keep values. In our solution, we use value as the key to insert and delete from the \textit{DynamicHeap} and keep value with its count (same value can be repeated) in the \textit{DynamicHeap}. In worst case it has O(log(n)) time complexity for both add and remove operations and in best case (i.e value already present and that does not cause median change) it has O(1) time complexity.

@@

The frequent route query entails finding the most frequent routes within the last 30 minutes. Algorithm \ref{frequent_route_algorithm} shows the algorithm to process this query using the \textit{NodeList} data structure and a queue to keep the events for the time window.  Since queue operations are supported in O(1) time, query evaluations happen in O(log(n)) time. Algorithm \ref{frequent_route_algorithm} shows the basic algorithm to evaluate the query an event. This algorithm assumes there is a global \textit{NodeList} object and route counts are stored in an object called \textit{RouteCount}.

@@

We have implemented profitable cells query in a similar manner to the frequent route query using  time windows. Complexity of the query raised following issues and we have solved them as given below.
\begin{enumerate}
	\item The profitability is calculated using the median profit and the number of empty taxis. In order to calculate median, we use \textit{java.util.PriorityQueue} based approach mentioned in (Section 2.4). We choose this option over our implementation, because this approach performs better with smaller window sizes. First, these heap sizes do not grow over 145 and there are same values repeating in the list increasing the probability of finding a value without traversing the whole list. Secondly, our implementation creates more objects per operation compared to just keeping double values though the time complexity is O(log(n)). In section 4.3 we show that our approach works better for larger window sizes.
	\item In order to keep track of empty taxis, we need to process both pickup cell and dropoff cell details for each message. One way of doing this is to send all events for the same process instance and process both details with one event. However this avoids the possibility of partitioning data across different processes. We generate two events, one for pickup data and one for dropoff data to make the process parallel. 
	\item In order to handle expiration of dropoff events, we keep track of the taxi ids currently available in a particular cell. If a pickup event occurs before dropoff event expiration, then we remove that taxi id from that cell and reduce the empty taxi count. In such a case, dropoff event expiration does not cause a reduction of taxi count.
\end{enumerate}

>>>>

\section{Query Implementation}
In this section, we elaborate how we have implemented the query logic to achieve O(log n) time complexity. The 2015 Grand Challenge problem has two queries. First is to evaluate top ten most frequent routes within the last 30 minutes and second is to evaluate top ten profitable areas within the last 15 minutes. We can use composite objects to manage details for each route and cell and compare them to find top 10 values. However, each time window can have thousands of route and cell details. Therefore, an efficient data structure is needed to add, remove, update and retrieve top ten values from a set of objects for a given time window efficiently. Since each event is received with route and cell details, the aforementioned operations should be performed using route and cell details as the key. In order to support all these functions with O(log n) time complexity we have developed a data structure called \textit{NodeList}. The remainder of this section describes how we have achieved this time complexity with the \textit{NodeList} data structure and how we implemented queries using that data structure.

\subsection{NodeList Data Structure}

\begin{table}
\centering
\caption{Operations of \textit{NodeList} Data Structure}
\begin{tabular}{|l|l|} \hline
Operation Signature & Time Complexity \\ \hline \hline
add(key : Object, value : NodeValue) & O(log n) \\ \hline 
get(key : Object) : NodeValue & O(1) \\ \hline
remove(key : Object) : NodeValue & O(log n) \\ \hline
decrementPosition(key : Object) & O(log n) \\ \hline
incrementPosition(key : Object) & O(log n) \\ \hline
getTopValues() : List<NodeValue> & O(1) \\ \hline
\end{tabular}
\label{nodelist_api}
\end{table}

Table \ref{nodelist_api} shows the operations of this data structure. The \textit{NodeValue} is an interface to plug any user defined type as a value. For an example, the route query can use a \textit{RouteCount} object to keep route count while profitable cells query can use \textit{CellProfit} object to keep track of fare details. This interface contains a method called \texttt{compare} to define the order of the values. The \textit{NodeList} data structure provides the notion of a sorted values starting from position 0 (highest value) to its users. The \texttt{decrementPostion} operation moves the corresponding value towards position 0 until the correct position according to sorted order. Therefore this method should be called after increasing the value of an existing object. Similarly \texttt{incrementPosition} operation can be used to move a value to correct position after decreasing the value of an existing object. \texttt{GetTopValues} operation returns the top ten values of the list. The query algorithm sections further elaborate on the usage of these methods.

One way of implementing this data structure is to use a heap. A heap supports adding and extracting maximum value operations in O(log n) time. However, since this application requires top 10 values, we need to retrieve maximum value ten times and insert them back. On the other hand if we use a doubly linked list to store values, it will take O(n) time for  \texttt{decrementPostion} and \texttt{incrementPosition} operations. Therefore we can use a doubly linked list to keep first 10 values (this makes time complexity for all operations with this data structure O(1)) and a heap to keep other values to avoid above problems.  A \textit{Map} which points to values can be used to retrieve values using a key.

@@

\subsection{OrderedList Data Structure}

@@

Table \ref{orderedlist_api} shows the operations of the \textit{OrderedList} data structure. First six methods have the same meaning as the \textit{NodeList} data structure and last two can be used to get the size of the list and the value of the last object. We internally keep a counter and a pointer to tail to support these operations in O(1) time. Figure \ref{orderedlist_impl} shows the implementation details. The \textit{Map} structure (an \textit{java.util.HashMap}) is used to retrieve object pointers with O(1) expected time. Each node of the doubly linked list points to previous and next nodes while storing the \textit{NodeValue}.

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{orderedlist.png}
        \caption {A schematic representation of the \textit{OrderedList} data structure}
        \label{orderedlist_impl}
\end{figure}

\subsection{DynamicHeap Data Structure}

\begin{table}
\centering
\caption{Operations of \textit{DynamicHeap} Data Structure}
\begin{tabular}{|l|l|} \hline
Operation Signature & Time Complexity \\ \hline \hline
add(key : Object, value : NodeValue) & O(log n) \\ \hline
remove(key : Object) : NodeValue & O(log n) \\ \hline
containsKey(key : Object) : boolean & O(1) \\ \hline
get(key : Object) : NodeValue & O(1) \\ \hline
moveUp(key : Object) & O(log n) \\ \hline
moveDown(key : Object) & O(log n) \\ \hline
getMaxkey() : Object & O(1) \\ \hline
extractMax() : NodeValue & O(log n) \\ \hline
\end{tabular}
\label{dynamicheap_api}
\end{table}

Table \ref{dynamicheap_api} shows the operations of the \textit{DynamicHeap} data structure. \texttt{MoveUp} and \texttt{moveDown} operations are used to move an element up in the heap if the value increased, or move down in the heap if the value decreased. Figure \ref{dynamicheap_impl} shows the implementation details. Conventional heaps only support \texttt{add} and \texttt{extractMax} operations since heap operations require knowing the position of the element in underlying array. In order to support \texttt{moveUp}, \texttt{moveDown} and \texttt{remove} operations, we keep the array index with the object itself. As shown in figure \ref{dynamicheap_impl} \textit{HeapNode} object contains the array index and the \textit{NodeValue} given by the user. This allows us to retrieve array index in O(1) time and move the element O(log n) time.

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{DinamicHeap.png}
        \caption {A schematic representation of the \textit{DynamicHeap} data structure}
        \label{dynamicheap_impl}
\end{figure}

\subsection{Finding the Runing Median}
In general, we can find the runing median using two heaps. One to keep values greater than the current median and one to keep values less than the current median. An implementation of this algorithm using two \textit{java.util.PriorityQueue}s have an O(log n) time complexity for add operation and O(n) time complexity for remove operation (\texttt{remove} operation of \textit{java.util.PriorityQueue} takes O(n) time).  We overcame this issue using \textit{DynamicHeap} to keep values. In our solution, we use value as the key to insert and delete from the \textit{DynamicHeap} and keep each value with its count (same value can be repeated) in the \textit{DynamicHeap}. In the worst case this has a O(log n) time complexity for both the add and remove operations and in the best case (i.e value is already present and it does not cause a median change) it has O(1) time complexity.

@@

The frequent route query entails finding the most frequent routes within the last 30 minutes. Algorithm \ref{frequent_route_algorithm} shows the algorithm to process this query using the \textit{NodeList} data structure and a queue to keep the events for the time window.  Since queue operations are supported in O(1) time, query evaluations happen in O(log n) time. Algorithm \ref{frequent_route_algorithm} shows the basic algorithm to evaluate the query an event. This algorithm assumes there is a global \textit{NodeList} object and route counts are stored in an object called \textit{RouteCount}.

@@

We have implemented profitable cells query in a similar manner to the frequent route query using  time windows. Complexity of the query raised following issues and we have solved them as given below.
\begin{enumerate}
	\item Profitability is calculated using the median profit and the number of empty taxis. In order to calculate the median, we use \textit{java.util.PriorityQueue} based approach mentioned in (Section 2.4). We choose this option over our implementation, because this approach performs better with smaller window sizes. First, these heap sizes do not grow over 145 and there are same values repeating in the list increasing the probability of finding a value without traversing the whole list. Secondly, our implementation creates more objects per operation compared to just keeping double values though the time complexity is O(log n). In section 4.3 we show that our approach works better for larger window sizes.
	\item In order to keep track of empty taxis, we need to process both pickup cell and dropoff cell details for each message. One way of doing this is to send all events for the same process instance and process both details with one event. However, this avoids the possibility of partitioning data across different processes. We generate two events, one for pickup data and one for dropoff data to make the process parallel. 
	\item In order to handle expiration of dropoff events, we keep track of the taxi ids currently available in a particular cell. If a pickup event occurs before dropoff event expiration, then we remove that taxi id from that cell and reduce the empty taxi count. In such a case, dropoff event expiration does not cause a reduction of taxi count.
\end{enumerate}
